---
title: "Assessment of Construction Segments"
output: html_document
date: "2024-03-17"
editor_options: 
  markdown: 
    wrap: 72
---

# Let's do this!

```{r}
# Environment
library(tidyverse)
library(lubridate)
```

## Inspect categories

-   Spend categories: Which sub-groups will we compare?

```{r}
# Load construction categories
summary_employees_and_earnings <- read_csv(
  "../datasets/generated_by_main/category_dataframe_0.csv")
summary_total_spend <- read_csv(
  "../datasets/generated_by_main/category_dataframe_1.csv")
summary_producer_price_index <- read_csv(
  "../datasets/generated_by_main/category_dataframe_2.csv")

# Identify category IDs to extract observational data
summary_total_spend %>% 
  filter(seasonal_adjustment_short == 'SAAR') %>% 
  filter(units == 'Millions of Dollars') %>% 
  filter(!grepl('Public|Private|in the United States', title)) %>% 
  select(id, title)
```

## Prepare Observation Data

With the IDs of the construction segments, I pulled the observation data
from the FRED API and prepared them as CSVs in Python. They will be
loaded here for analysis.

```{r}
# Bulk load observation files
csv_files <- list.files(path = "../datasets/generated_by_main/", 
                        pattern = "^observation_dataframe_.*\\.csv$")
list_of_dfs <- lapply(
  csv_files, 
  function(file) read.csv(file.path("../datasets/generated_by_main/", file)))

# Bind all construction segments
df_total_spend <- bind_rows(list_of_dfs)

# Format total spend data
df_total_spend <- df_total_spend %>% 
  mutate(description = str_replace(description, "Total Construction Spending: ", ""),
         year = year(date)) %>% 
  filter(realtime_end == '9999-12-31') %>% 
  select(code, description, date, year, value)

# Set current year
current_year <- year(Sys.Date())

# Remove files from env
rm(csv_files)
rm(list_of_dfs)
```

## Exploration

```{r}
# Date range of available data
df_total_spend %>% 
  group_by(description) %>%
  summarise(earliest_date = min(date),
            latest_date = max(date))

# Calculate total value for varying N years for all segments
df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 3)) %>%
  summarise(total_3yr = sum(value)) %>% 
  arrange(desc(total_3yr))

df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 5)) %>%
  summarise(total_5yr = sum(value)) %>% 
  arrange(desc(total_5yr))

df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 10)) %>%
  summarise(total_10yr = sum(value)) %>% 
  arrange(desc(total_10yr))

df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 20)) %>%
  summarise(total_20yr = sum(value)) %>% 
  arrange(desc(total_20yr))
```

-   We have at least 22 years of data for all segments.
-   Next steps are to determine metrics for size and growth.

For *size*, I want a look back period of 5 years to understand current
market dynamics. I will also look at 10 years to capture a broader view
of long-term market share changes and trends.

For *growth*, I think the last 3 years are ideal for seeing emerging
growth trends and identifying sectors are gaining momentum. However, I
will also look at the last 5 year window which serves as a common
reference point in the narrative.

```{r}
# Aggregate segment spend, minus overlapping categories
overall_total_5yr <- df_total_spend %>%
  filter(year >= (current_year - 5), 
         !description %in% c("Total Construction Spending", 
                             "Nonresidential")
         ) %>%
  summarise(total_5yr = sum(value)) %>%
  pull(total_5yr)

overall_total_10yr <- df_total_spend %>%
  filter(year >= (current_year - 10), 
         !description %in% c("Total Construction Spending", "Nonresidential")) %>%
  summarise(total_10yr = sum(value)) %>%
  pull(total_10yr)

# Construct metrics for size and growth
df_size_growth <- df_total_spend %>%
  group_by(description) %>%
  summarise(
    # Size calculations: need `overall_total_*_yr` for share of construction
    total_value_5yr = sum(value[year >= (current_year - 5)]),
    total_value_10yr = sum(value[year >= (current_year - 10)]),
    size_percent_5yr = (total_value_5yr / overall_total_5yr) * 100,
    size_percent_10yr = (total_value_10yr / overall_total_10yr) * 100,
    # Growth calculations: need `start` + `end` values for each segment for CAGR
    start_value_3yr = first(value[year == (current_year - 3)]),
    end_value_3yr = last(value[year == current_year]),
    start_value_5yr = first(value[year == (current_year - 5)]),
    end_value_5yr = last(value[year == current_year]),
    CAGR_3yr = if_else(end_value_3yr != 0 & start_value_3yr != 0, 
                       ((end_value_3yr/start_value_3yr)^(1/3) - 1) * 100, NA_real_),
    CAGR_5yr = if_else(end_value_5yr != 0 & start_value_5yr != 0, 
                       ((end_value_5yr/start_value_5yr)^(1/5) - 1) * 100, NA_real_)
  ) %>%
  filter(!is.na(CAGR_3yr), 
         !is.na(CAGR_5yr), 
         !description %in% c("Total Construction Spending", "Nonresidential")) %>% 
  mutate_if(is.numeric, round, 2)

# Format summary table, by construction segment
(
  segment_summary <- df_size_growth %>% 
    select(segment = description,
           market_share_5yr = size_percent_5yr,
           market_share_10yr = size_percent_10yr,
           growth_rate_3yr = CAGR_3yr,
           growth_rate_5yr = CAGR_5yr) %>% 
    arrange(desc(market_share_5yr))
  )
```
## ANALYSIS PLAN

0.  Preparation
    -   Filter to earliest common date across all segments
    -   Annualize numbers
    -   Code: Residential vs Non-Residential vs NA
1.  Size, %: Who are the largest segments in construction? How big is
    residential? Who are the biggest in non-residential, outside of
    commercial?
2.  Growth, %: Who is growing faster than the average?
3.  Labor, \# of employees: Are any segments facing greater deficits of
    workers, relative to spend (make assumptions)? And are mean wages
    declining faster in some segments? Might need to recode or remove
    groups.
4.  Materials, PPI: Are costs for sub-contractors rising in some
    segments faster than others?

Looking for segments with higher size, higher growth, lower employees,
and higher materials.

## SCRATCH

PPI - Producer Price Index by Industry: New Office Building Construction
