---
title: "Construction Market Analysis"
output: html_document
date: "2024-03-17"
author: "Tim Lu"
editor_options: 
  markdown: 
    wrap: 72
---

## Preparation

```{r}
# Load libraries
library(tidyverse)
library(lubridate)
library(tools)
```

```{r}
# Load construction categories from CSV files
summary_employees_and_earnings <- read_csv("../datasets/generated_by_main/category_dataframe_0.csv")
summary_total_spend <- read_csv("../datasets/generated_by_main/category_dataframe_1.csv")
summary_producer_price_index <- read_csv("../datasets/generated_by_main/category_dataframe_2.csv")

# Identify relevant category IDs for further analysis, excluding broad aggregates
summary_total_spend %>% 
  filter(seasonal_adjustment_short == 'SAAR') %>% 
  filter(units == 'Millions of Dollars') %>% 
  filter(!grepl('Public|Private|in the United States', title)) %>% 
  select(id, title)
```

-   With the above IDs for segments of interest, I pulled the relevant
    observation data from the FRED API and transformed them into CSVs
    (Python).

-   Here, I load those CSVs with economic data across segments to
    prepare analysis.

```{r message=FALSE}
# Bulk load observation files
csv_files <- list.files(
  path = "../datasets/generated_by_main/", 
  pattern = "^observation_dataframe_.*\\.csv$"
  )
list_of_dfs <- lapply(
  X = csv_files, 
  FUN = function(file) read_csv(file.path("../datasets/generated_by_main/", file))
  )

# Bind data for all construction segments
df_total_spend <- bind_rows(list_of_dfs)

# Format the total spend data
df_total_spend <- df_total_spend %>% 
  mutate(description = str_replace(description, "Total Construction Spending: ", ""),
         year = year(date)) %>% 
  filter(realtime_end == '9999-12-31') %>% 
  select(code, description, date, year, value)

# Set current year for time-based filtering later
current_year <- year(Sys.Date())

# Cleanup: Remove temp variables from environment
rm(list = c("csv_files", "list_of_dfs"))
```

## Exploratory Data Analysis (EDA)

```{r}
# Date range of available data
df_total_spend %>% 
  group_by(description) %>%
  summarise(earliest_date = min(date),
            latest_date = max(date))

# Calculate total value for varying N years for all segments
df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 3)) %>%
  summarise(total_3yr = sum(value)) %>% 
  arrange(desc(total_3yr))

df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 5)) %>%
  summarise(total_5yr = sum(value)) %>% 
  arrange(desc(total_5yr))

df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 10)) %>%
  summarise(total_10yr = sum(value)) %>% 
  arrange(desc(total_10yr))

df_total_spend %>% 
  group_by(description) %>%
  filter(year >= (current_year - 20)) %>%
  summarise(total_20yr = sum(value)) %>% 
  arrange(desc(total_20yr))
```

-   We have at least 22 years of data for all segments.
-   Next steps are to determine metrics for size and growth.
    -   For *size*, I want a look back period of 5 years to understand
        current market dynamics. I will also look at 10 years to capture
        a broader view of long-term market share changes and trends.
    -   For *growth*, I think the last 3 years are ideal for seeing
        emerging growth trends and identifying sectors are gaining
        momentum. However, I will also look at the last 5 year window
        which serves as a common reference point (together with size /
        market share) in the narrative.

```{r}
df_total_spend %>%
  filter(year >= (current_year - 8),
         description %in% c("Total Construction Spending", "Nonresidential", "Residential")
         ) %>%
  mutate(value = value / 1000000,
         description = ifelse(description == "Total Construction Spending",
                              "Total", description)) %>% 
  ggplot(aes(x = year, y = value, color = description, group = description, linetype = description)) +
  geom_smooth(se = FALSE, method = "loess") +
  theme_minimal() +
  labs(
    y = "Share of Construction Spend ($ Trillion)",
    x = "",
    color = "Segment", linetype = "Segment"
  ) +
  scale_color_manual(values = c(
    "Total" = "black",
    "Nonresidential" = "black",
    "Residential" = "#17A6F7"
  )) +
  scale_linetype_manual(values = c(
    "Total" = "dashed",
    "Nonresidential" = "solid",
    "Residential" = "solid"
  )) +
  theme(legend.position = "bottom") 
```

## Metrics Construction

```{r}
# Determine categories to exclude for analysis
category_blacklist <- c("Total Construction Spending",
                        "Nonresidential" # Comment out if including `Nonresidential`
                        # "Residential"    # Comment out if including `Residential`
                        )

# Aggregate segment spend, minus overlapping categories
overall_total_5yr <- df_total_spend %>%
  filter(year >= (current_year - 5), 
         !description %in% category_blacklist
         ) %>%
  summarise(total_5yr = sum(value)) %>%
  pull(total_5yr)

overall_total_10yr <- df_total_spend %>%
  filter(year >= (current_year - 10), 
         !description %in% category_blacklist
         ) %>%
  summarise(total_10yr = sum(value)) %>%
  pull(total_10yr)

# Construct metrics for size and growth
df_size_growth <- df_total_spend %>%
  group_by(description) %>%
  summarise(
    # Size calculations: need `overall_total_*_yr` for share of construction
    total_value_5yr = sum(value[year >= (current_year - 5)]),
    total_value_10yr = sum(value[year >= (current_year - 10)]),
    size_percent_5yr = (total_value_5yr / overall_total_5yr) * 100,
    size_percent_10yr = (total_value_10yr / overall_total_10yr) * 100,
    # Growth calculations: need `start` + `end` values for each segment for CAGR
    start_value_3yr = first(value[year == (current_year - 3)]),
    end_value_3yr = last(value[year == current_year]),
    start_value_5yr = first(value[year == (current_year - 5)]),
    end_value_5yr = last(value[year == current_year]),
    CAGR_3yr = if_else(end_value_3yr != 0 & start_value_3yr != 0, 
                       ((end_value_3yr/start_value_3yr)^(1/3) - 1) * 100, NA_real_),
    CAGR_5yr = if_else(end_value_5yr != 0 & start_value_5yr != 0, 
                       ((end_value_5yr/start_value_5yr)^(1/5) - 1) * 100, NA_real_)
  ) %>%
  filter(!is.na(CAGR_3yr), 
         !is.na(CAGR_5yr), 
         !description %in% category_blacklist
         ) %>% 
  mutate_if(is.numeric, round, 2)

# Format summary table, by construction segment
(
  segment_summary <- df_size_growth %>% 
    select(segment = description,
           market_share_5yr = size_percent_5yr,
           market_share_10yr = size_percent_10yr,
           growth_rate_3yr = CAGR_3yr,
           growth_rate_5yr = CAGR_5yr) %>% 
    arrange(desc(market_share_5yr))
  )

# Sanity checks
sum(segment_summary$market_share_5yr)
sum(segment_summary$market_share_10yr)
```

## Analysis Functions

```{r}
########################
## PREPARE PLOT TABLE ##
########################
prepare_plot_table <- function(df, metric_name, dimension_name, filter_field = NULL) {
  if (!is.null(filter_field)) {
    df <- df %>% filter(segment != filter_field)
  }
  
  mean_metric <- mean(df[[metric_name]], na.rm = TRUE)
  sd_metric <- sd(df[[metric_name]], na.rm = TRUE)
  
  df <- df %>%
    mutate(
      color_group = case_when(
        .data[[metric_name]] > mean_metric + sd_metric ~ "above_1_sd",
        .data[[metric_name]] < mean_metric - sd_metric ~ "below_1_sd",
        TRUE ~ "between"
      ),
      !!sym(dimension_name) := forcats::fct_reorder(.data[[dimension_name]],
                                                    .data[[metric_name]], 
                                                    .desc = FALSE)
    )
  
  return(df)
}

###############################
## CREATE HORIZONTAL BARPLOT ##
###############################
create_horizontal_barplot <- function(df,
                                      metric_name,
                                      dimension_name,
                                      grouping_var = NULL
                                      ) {

  p <- ggplot(df, aes(x = .data[[dimension_name]],
                      y = .data[[metric_name]],
                      fill = color_group)) +
    geom_bar(stat = 'identity') +
    geom_text(aes(label = sprintf("%.1f", .data[[metric_name]]),
                  y = .data[[metric_name]]),
              position = position_dodge(width = 0.9),
              hjust = -0.2, size = 3.5, color = "grey60") +
    scale_x_discrete(labels = function(x) str_wrap(x, width = 80)) +
    scale_fill_manual(values = c("above_1_sd" = "#17A6F7FF",
                                 "below_1_sd" = "#17A6F766",
                                 "between" = "gray")) +
    coord_flip() +
    labs(
      x = toTitleCase(gsub("_", " ", dimension_name)),
      y = toTitleCase(gsub("_", " ", metric_name))
    ) +
    theme_minimal() +
    theme(legend.position = "none", text = element_text(size = 12, color="grey20"))

  if (!is.null(grouping_var)) {
    p <- p + facet_wrap(as.formula(paste("~", grouping_var)))
  }

  return(p)
}
```

## Market Share Analysis

```{r}
# Including `Residential`
prepare_plot_table(df = segment_summary, 
                   metric_name = "market_share_5yr", 
                   dimension_name = "segment") %>% 
  create_horizontal_barplot(df = .,
                   metric_name = "market_share_5yr",
                   dimension_name = "segment")

# Including `Residential`
prepare_plot_table(df = segment_summary, 
                   metric_name = "growth_rate_5yr", 
                   dimension_name = "segment") %>% 
  mutate(color_group = ifelse(segment == 'Residential', "above_1_sd", "between")) %>% 
  create_horizontal_barplot(df = .,
                   metric_name = "growth_rate_5yr",
                   dimension_name = "segment")
```

```{r}
# Excluding `Residential`
# Window: 5 Year
table_market_share_5yr <- prepare_plot_table(df = segment_summary, 
                                             metric_name = "market_share_5yr", 
                                             dimension_name = "segment", 
                                             filter_field = "Residential")

(
  plot_market_share_5yr <- create_horizontal_barplot(df = table_market_share_5yr,
                                                   metric_name = "market_share_5yr",
                                                   dimension_name = "segment")
  )

# Window: 10 Year
table_market_share_10yr <- prepare_plot_table(df = segment_summary, 
                                             metric_name = "market_share_10yr", 
                                             dimension_name = "segment", 
                                             filter_field = "Residential")

(
  plot_market_share_10yr <- create_horizontal_barplot(df = table_market_share_10yr,
                                                   metric_name = "market_share_10yr",
                                                   dimension_name = "segment")
  )
```

## Growth Trend Analysis

```{r}
# Excluding `Residential`
# Window: 3 Year
table_growth_rate_3yr <- prepare_plot_table(df = segment_summary, 
                                            metric_name = "growth_rate_3yr", 
                                            dimension_name = "segment",
                                            filter_field = "Residential")

(
  plot_growth_rate_3yr <- create_horizontal_barplot(df = table_growth_rate_3yr,
                                                   metric_name = "growth_rate_3yr",
                                                   dimension_name = "segment")
  )

# Window: 5 Year
table_growth_rate_5yr <- prepare_plot_table(df = segment_summary, 
                                             metric_name = "growth_rate_5yr", 
                                             dimension_name = "segment",
                                            filter_field = "Residential")

(
  plot_growth_rate_5yr <- create_horizontal_barplot(df = table_growth_rate_5yr,
                                                   metric_name = "growth_rate_5yr",
                                                   dimension_name = "segment")
  )
```

## Material Costs and Price Index Analysis

<https://fred.stlouisfed.org/release/tables?rid=46&eid=134088>

-   A high PPI indicates:
    -   (+) selling prices for new industrial building construction are
        increasing
    -   (+) which means this sub-segment is healthy
    -   (+) and strong demand allowing builders and subcontractors to
        charge more
    -   (+) thus, potentially higher budget allocations towards our SaaS
-   Gonna assume that `PPI for New Industrial Building Construction` is
    reasonable proxy of the broader manufacturing segment's price
    changes over time.

```{r}
# To save time, I'll input 5yr `start` + `end` values for PPI manually
construction_data <- data.frame(
  new_building_construction = c(
    "Industrial", # https://fred.stlouisfed.org/series/PCU236211236211
    "Warehouse",  # https://fred.stlouisfed.org/series/PCU236221236221
    "School",     # https://fred.stlouisfed.org/series/PCU236222236222
    "Office",     # https://fred.stlouisfed.org/series/PCU236223236223
    "Health Care" # https://fred.stlouisfed.org/series/PCU236224236224
    ),
  start = c(132.7, 155.5, 165.9, 140.6, 115.3),
  end = c(193.3, 232.3, 224.5, 205.1, 157.9)
)

ppi_summary <- construction_data %>%
  mutate(percent_change_in_producer_price_index_5yr = ((end - start) / start) * 100) %>% 
  arrange(desc(percent_change_in_producer_price_index_5yr))

# Window: 5 year
table_ppi_5yr <- prepare_plot_table(
  df = ppi_summary, 
  metric_name = "percent_change_in_producer_price_index_5yr", 
  dimension_name = "new_building_construction") 

(
  plot_ppi_5yr <- create_horizontal_barplot(
    df = table_ppi_5yr,
    metric_name = "percent_change_in_producer_price_index_5yr",
    dimension_name = "new_building_construction")
  )
```

## Labor Market Dynamics

-   It seems that employee count and wage data is only available at the
    residential vs non-residential level.
    <https://fred.stlouisfed.org/release/tables?rid=50&eid=4881#snid=4899>

-   Since I already know Residential is less appealing than
    Non-Residential's Manufacturing sub-segment, I will forgo this part
    on labor market dynamics.

# APPENDIX

## Analysis Plan

0.  Preparation
    -   Filter to earliest common date across all segments
    -   Annualize numbers
    -   Bin Residential vs Non-Residential vs NA
1.  Size, %: Who are the largest segments in construction? How big is
    residential? Who are the biggest in non-residential, outside of
    commercial?
2.  Growth, %: Who is growing faster than the average?
3.  Labor, \# of employees: Are any segments facing greater deficits of
    workers, relative to spend (make assumptions)? And are mean wages
    declining faster in some segments? Might need to recode or remove
    groups.
4.  Materials, PPI: Are costs for sub-contractors rising in some
    segments faster than others?

Looking for segments with higher size, higher growth, lower employees,
and higher materials.
